Notes on thermodnyamic relations in {\tt MAESTRO}

### 2010-08-11 docs/thermo_notes/thermo_notes.tex
add a simple derivation of the adiabatic excess using quantities returnable from our EOS

>>>>

\section{Adiabatic Excess}
\label{Sec:Adiabatic Excess}
The adiabatic excess, $\Delta\nabla$, is a quantity used to determine
if a system is stable ($\Delta\nabla < 0$) or unstable ($\Delta\nabla
> 0$) to convection under the Schwarzschild criterion (i.e. neglecting
compositional gradients).  Cox and Giuli (see chapter 9) define three
different ``adiabatic exponents'' that we will use:
\begin{eqnarray*}
  \Gamma_1 &\equiv&   \left(\frac{d\ln p}{d\ln\rho}\right)_\text{ad} \\
  \frac{\Gamma_2}{\Gamma_2-1} &\equiv& 
  \left(\frac{d\ln p}{d\ln T}\right)_\text{ad} \\
  \Gamma_3 - 1 &\equiv& \left(\frac{d\ln T}{d\ln\rho}\right)_\text{ad},
\end{eqnarray*} 
where the subscript ``ad'' means along an adiabat.  We can combine the
exponents to get the following relation
\begin{equation}\label{eq:Gamma relations}
  \Gamma_1 = \left(\frac{\Gamma_2}{\Gamma_2-1}\right)\left(\Gamma_3-1\right).
\end{equation}

The adiabatic excess is defined as
\begin{equation}\label{eq:adiabatic excess}
  \Delta\nabla = \nabla_\text{actual} - \nabla_\text{ad}
\end{equation}
where
\begin{equation}\label{eq:thermal gradient}
  \nabla \equiv \frac{d\ln T}{d\ln P}
\end{equation}
is the thermal gradient.  It is important to note that these thermal
gradients are only along the radial direction.  The ``actual''
gradient can be found from finite differencing the data whereas the
adiabatic term, $\nabla_\text{ad} = \left(\Gamma_2-1\right) /
\Gamma_2$, will need to be calculated at each point using
thermodynamic relations.  Our EOS only returns $\Gamma_1$ so we need
find another relation to use with \eqref{eq:Gamma relations} to solve
for the adiabatic excess.

The Schwarzschild criterion does not care about changes in composition
and we therefore write $p = p(\rho,T)$ and
\begin{equation}\label{eq:dp}
  d\ln p = \chi_\rho d\ln\rho + \chi_T d\ln T
\end{equation}
where
\[
\chi_\rho = \left(\frac{d\ln p}{d\ln\rho}\right)_T,\qquad
\chi_T = \left(\frac{d\ln p}{d\ln T}\right)_\rho.
\]
Dividing \eqref{eq:dp} by $d\ln\rho$ and taking this along an adiabat
we have
\begin{equation}\label{eq:dp2}
  \left(\frac{d\ln p}{d\ln\rho}\right)_\text{ad} = \chi_\rho + \chi_T
  \left(\frac{d\ln T}{d\ln\rho}\right)_\text{ad}.
\end{equation}
Using the $\Gamma$'s, we have
\begin{equation}\label{eq:Gamma1 relation with Gamma2}
  \Gamma_1 = \chi_\rho + \chi_T\left(\Gamma_3-1\right).
\end{equation}
Combining \eqref{eq:Gamma relations} and \eqref{eq:Gamma1 relation
  with Gamma2} to eliminate $\Gamma_3$, we have:
\begin{equation}\label{eq:nabla_ad}
  \nabla_\text{ad} = \frac{\Gamma_1 - \chi_\rho}{\chi_T\Gamma_1}
\end{equation}
which uses only terms which are easily returned from an EOS call.
